The Two Dualities of Computation: Negative and Fractional Types

### 2012-03-11 rational.tex
reorg
sec negative types

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Negative Types}
\label{sec:neg}

The structure of this section and the next are the same. The general outline
is as follows:\footnote{We refer the reader to Selinger's excellent survey
  article of monoidal
  categories~\cite{springerlink:10.1007/978-3-642-12821-94} for the precise
  definitions.}
\begin{itemize}
\item As established in our earlier work~\cite{rc2011,infeffects}, the
  underlying categorical semantics of our core reversible language
  {{langRev}} is based on \emph{two} distinct symmetric monoidal structures,
  one with {{+}} as the monoidal tensor, and one with {{*}} as the monoidal
  tensor.
\item We extend each underlying symmetric monoidal structure to a
  \emph{compact closed} structure by adding a dual for each object and two
  special morphims traditionally called {{eta}} and {{eps}}. 
\end{itemize}
After presenting the extension at the syntactic level, we discuss the
categorical semantics informally via the graphical language. The operational
semantics is presented in Sec.~\ref{sec:rat}. As will be detailed in this
section and the next, the compact closed structure provides several
properties of interest: (i) morphisms or wires are allowed to run from right
to left, (ii) the structure admits a trace that, depending on the situation,
can be used to implement various notions of loops and
recursion~\cite{joyal1996traced,Hasegawa:2009:TMC:1552068.1552069,Hasegawa:1997:RCS:645893.671607},
(iii) the structure includes an isomorphism showing that the dual operator is
an involution, and (iv) the structure is equipped with objects representing
higher-order functions and a morphism \textit{eval} that applies these
functional objects. Interestingly each monoidal structure provides the same
ingredients, resulting in two dualities, two traces, two involutions, and two
notions of higher-order functions.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Additive Duality in {{langRev}} }

@@

\paragraph*{Negative Information Flow.} 
Hence, the intuitive way to read the diagrams is that they reverse the
flow of information -- in normal {{langRev}} circuits information
flows from left-to-right and {{eps+}} operations cause information to
flow from right-to-left.  The interpretation of {{eta+}} is precisely
the opposite: it turns a right-to-left flow of information in a
circuit into a left-to-right flow.
As an example, the diagram representing the isomorphism from the
introduction is:
\begin{center}
  \includegraphics{diagrams/algebra1.pdf}
\end{center}

\paragraph*{Time Traveling.}
A time traveling intuition is also applicable here: if view the
left-to-right direction of information flow as the canonical forward
direction of information, then the action of {{eps+}} allows values,
aka information particles, to flow backwards in time. Particles that
flow backwards in time get to see and interact with the history of
particles that they coexist with i.e. values that they exist
multiplicatively with. Hence the isomorphism {{(-b1)*b2<->-(b1*b2)}}
(whose construction is shown in Section \ref{sec:neg-constructions}).

%\subsection{Compact Closure with Additive Inverses}
\paragraph*{Coherence for Compact Closure with Additive Inverses.}
The appropriate coherence condition for Compact Closed Categories,
expressed as the triangle axiom, is satisfied by {{eta+}}/{{eps+}} and
may be visualized as explained by Selinger
\cite{springerlink:10.1007/978-3-642-12821-94} as follows:

@@

\paragraph*{Non-termination.}
As is well known in category theory, Compact Closed Categories contain
trace operations.  In the context of {{langRev}} this implies that
{{eta+}}/{{eps+}} allow the construction of a {{trace+}} operator
(detailed in Section \ref{sec:monoidal-constructions}) hence
introducing \emph{non-termination} into the language (in the presence
of recursive types) \cite{infeffects}.

\paragraph*{Zero Information.}
The fact that {{eta+}} (and hence {{eps+}}) introduces no information
can be understood is by viewing {{-b+b}} as a type whose value is
observable only if you `pay' {{b}} amount of information to the {{-b}}
branch. When {{b}} amount of information is supplied {{(-b+b)+b=b}}
amount of information is returned to you. (TODO. more needs to be said
here.)

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Fractional Types}
\label{sec:frac}

@@

%%%%%%%%%%%%%%%%%%%%%%
\subsection{Multiplicative Duality in {{langRev}} }

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Constructions with Negatives and Fractions}

@@

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Negative and Fractional Types}

% The structure of this section and the next are the same.

This section introduces the syntax and graphical languages for
negative types and fractionals.The general outline is as follows:
\footnote{We refer the reader to Selinger's excellent survey
  article of monoidal
  categories~\cite{springerlink:10.1007/978-3-642-12821-94} for the precise
  definitions.}
\begin{itemize}
\item As established in our earlier work~\cite{rc2011,infeffects}, the
  underlying categorical semantics of our core reversible language
  {{langRev}} is based on \emph{two} distinct symmetric monoidal structures,
  one with {{+}} as the monoidal tensor, and one with {{*}} as the monoidal
  tensor.
\item We extend each underlying symmetric monoidal structure to a
  \emph{compact closed} structure by adding a dual for each object and two
  special morphims traditionally called {{eta}} and {{eps}}. 
\end{itemize}
After presenting the extension at the syntactic level, we discuss the
categorical semantics informally via the graphical language. The
operational semantics is presented in Sec.~\ref{sec:rat}. As will be
detailed in this sections \ref{sec:neg} and \ref{sec:frac}, the
compact closed structure provides several properties of interest: (i)
morphisms or wires are allowed to run from right to left, (ii) the
structure admits a trace that, depending on the situation, can be used
to implement various notions of loops and
recursion~\cite{joyal1996traced,Hasegawa:2009:TMC:1552068.1552069,Hasegawa:1997:RCS:645893.671607},
(iii) the structure includes an isomorphism showing that the dual
operator is an involution, and (iv) the structure is equipped with
objects representing higher-order functions and a morphism
\textit{eval} that applies these functional objects. Interestingly
each monoidal structure provides the same ingredients, resulting in
two dualities, two traces, two involutions, and two notions of
higher-order functions.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Additive Duality in {{langRev}} }
\label{sec:neg}

@@

Hence, the intuitive way to read the diagrams is that they reverse the
flow of information -- i.e. they capture a notion of \emph{negative
  information flow}. In normal {{langRev}} circuits, information flows
from left-to-right. The operator {{eps+}} causes information to flow
from right-to-left.  The interpretation of {{eta+}} is the opposite:
it turns a right-to-left flow of information in a circuit into a
left-to-right flow.  The diagram below is an example of using {{eta+}}
and {{eps+}} and corresponds to the additive {{a <-> a}} isomorphism
from the introduction. (Here the {{0}} wires have been shown for
completeness.)
\begin{center}
\scalebox{1.2}{
  \includegraphics{diagrams/algebra1.pdf}
}
\end{center}

% \paragraph*{Time Traveling.}
A \emph{time traveling} intuition is also applicable here.  In a
normal cicruit, as computational steps are taken values flow from left
to right, i.e. computational time progresses from left to right.  The
action of {{eps+}} causes values, aka information particles, to flow
from right to left i.e. backwards in computational time.  The
interesting thing about this interpretation is that particles that
travel backwards in time get to see and interact with the history of
particles that they coexist with. This gives us an intuitive
interpretation of the isomorphism {{(-b1)*b2<->-(b1*b2)}} (see
Sec. \ref{sec:specific-constructions}) where it would otherwise seem
that {{-b1}} and {{b2}} move in opposite directions. The backward flow
of the {{-b1}} allows it to `see the past' of {{b2}} and is thus
equivalent to both particles moving backward in time.

%\subsection{Compact Closure with Additive Inverses}
\emph{Coherence for Compact Closure with Additive Duality.}  The
appropriate coherence condition for Compact Closed Categories,
expressed as the triangle axiom, is satisfied by {{eta+}}/{{eps+}} and
may be visualized (modulo {{0}} wires, swapping etc.) as explained by
Selinger \cite{springerlink:10.1007/978-3-642-12821-94} as the
operational equivalence:

@@

\emph{Non-termination.}  Compact closed categories contain {{trace}}
operations and this implies that {{eta+}}/{{eps+}} allow the
construction of a {{trace+}} operator (shown in Section
\ref{sec:monoidal-constructions}).  As shown in previous work
\cite{infeffects}, non-terminating computation can be constructed from
     {{trace+}} and recursive types.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \section{Fractional Types}

%%%%%%%%%%%%%%%%%%%%%%
\subsection{Multiplicative Duality in {{langRev}} }
\label{sec:frac}

@@

@@

@@

\emph{Zero Information.}  The intuitive explanation that {{eta+}} and
hence {{eps+}} introduces no information effects is that {{-b+b}} as a
type whose value is observable only if you `pay' {{b}} amount of
information to the {{-b}} branch. When {{b}} amount of information is
supplied {{(-b+b)+b=b}} amount of information is returned to
you. (TODO. more needs to be said here.)
